---
title: "2D ENS Differentiation Protocol"
subtitle: "Studer / Fattahi Option B — Barber et al. 2019 · Majd et al. 2025"
author: "JP Flores"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    embed-resources: true
    css: print.css
prefer-html: true
server: shiny
---

```{r setup}
#| include: false
library(shiny)
library(tidyverse)
library(knitr)
library(kableExtra)

fmt <- function(x, d = 1) formatC(round(x, d), format = "f", digits = d)
ex  <- function(n, p = 0.10) n * (1 + p)
CM2 <- 9.5

cocktail_a <- function(w) {
  vol <- w * CM2 * 0.2
  tibble(
    Component     = c("Essential 6 (E6)", "BMP4 (10 µg/mL)",
                      "SB431542 (10 mM)", "CHIR 99021 (6 mM)"),
    Catalog       = c("Life Tech A1516401", "R&D 314-BP", "R&D 1614", "Tocris 4423"),
    `Final conc.` = c("base", "1 ng/mL", "10 µM", "600 nM"),
    `Amount (µL)` = fmt(c(vol*(1-1e-4-1e-3-1e-4), vol*1e-4, vol*1e-3, vol*1e-4))
  )
}
cocktail_b <- function(w) {
  vol <- w * CM2 * 0.2
  tibble(
    Component     = c("Essential 6 (E6)", "SB431542 (10 mM)", "CHIR 99021 (6 mM)"),
    `Final conc.` = c("base", "10 µM", "1.5 µM"),
    `Amount (µL)` = fmt(c(vol*(1-1e-3-2.5e-4), vol*1e-3, vol*2.5e-4))
  )
}
cocktail_c <- function(w) {
  v6 <- w*CM2*0.4; v10 <- w*CM2*0.6
  tibble(
    Component       = c("Essential 6 (E6)", "SB431542 (10 mM)",
                        "CHIR 99021 (6 mM)", "Retinoic Acid (1 mM)"),
    `Final conc.`   = c("base","10 µM","1.5 µM","1 µM"),
    `Days 6-8 (µL)` = fmt(c(v6*(1-1e-3-2.5e-4-1e-3), v6*1e-3, v6*2.5e-4, v6*1e-3)),
    `Day 10 (µL)`   = fmt(c(v10*(1-1e-3-2.5e-4-1e-3), v10*1e-3, v10*2.5e-4, v10*1e-3))
  )
}
ncc_medium <- function(w) {
  vol <- w * 2
  tibble(
    Component    = c("Neurobasal medium","N2 (100×)","B27 (50×)",
                     "Glutagro (200 mM)","MEM NEAAs (100×)",
                     "FGF2 (10 µg/mL)","CHIR 99021 (6 mM)"),
    Catalog      = c("Life Tech 21103049","CTS A1370701","Life Tech 17504044",
                     "Corning 25-015-CI","Corning 25-025-CI","R&D 233-FB","Tocris 4423"),
    `Per mL`     = c("base","10 µL","20 µL","10 µL","10 µL",
                     "1 µL → 10 ng/mL","0.5 µL → 3 µM"),
    `Total (µL)` = fmt(c(vol*1e3*(1-0.01-0.02-0.01-0.01-0.001-5e-4),
                          vol*10,vol*20,vol*10,vol*10,vol*1,vol*0.5))
  )
}
enc_medium <- function(w) {
  vol <- w * CM2 * 0.2
  tibble(
    Component       = c("Neurobasal medium","N2 (100×)","B27 (50×)",
                        "Glutagro (200 mM)","MEM NEAAs (100×)",
                        "GDNF (10 µg/mL)","Ascorbic Acid (200 mM)"),
    Catalog         = c("Life Tech 21103049","CTS A1370701","Life Tech 17504044",
                        "Corning 25-015-CI","Corning 25-025-CI",
                        "Peprotech 450-10","Sigma"),
    `Per mL`        = c("base","10 µL","20 µL","10 µL","10 µL",
                        "1 µL → 10 ng/mL","0.5 µL → 100 µM"),
    `Per feed (µL)` = fmt(c(vol*(1-0.01-0.02-0.01-0.01-0.001-5e-4),
                              vol*0.01,vol*0.02,vol*0.01,vol*0.01,vol*0.001,vol*5e-4))
  )
}
coating_tbl <- function(w) {
  vol <- w * CM2 * 0.1
  tribble(
    ~Step,                           ~Reagent,                                                    ~Temp,   ~Time,
    "1 — PLO overnight",             paste0("PLO 15 µg/mL in borate buffer — ",fmt(vol,0)," µL"), "37°C",  "O/N",
    "2 — Wash",                      "PBS × 3",                                                    "RT",    "—",
    "3 — LM + FN (≥2 hr)",          paste0("Laminin 2 µg/mL + FN 2 µg/mL in PBS — ",fmt(vol,0)," µL"),"37°C","≥2 hr",
    "4 — Aspirate before plating",   "Do NOT let surface dry",                                     "—",     "—"
  )
}
day_schedule <- function(w) {
  cm2 <- w * CM2
  tribble(
    ~Day,    ~Medium,       ~`µL/cm²`, ~`Total (µL)`,      ~Notes,
    "0",     "Cocktail A",  "200",     fmt(cm2*0.2,0),     "Switch from E8; record Day 0",
    "2",     "Cocktail B",  "200",     fmt(cm2*0.2,0),     "—",
    "4",     "Cocktail B",  "200",     fmt(cm2*0.2,0),     "Full medium change",
    "6",     "Cocktail C",  "400",     fmt(cm2*0.4,0),     "⚠ Volume increases; add RA",
    "8",     "Cocktail C",  "400",     fmt(cm2*0.4,0),     "Tilt plate; add from corner",
    "10",    "Cocktail C",  "600",     fmt(cm2*0.6,0),     "⚠ Volume increases again",
    "11-12", "Harvest →",   "—",       "—",                "QC first; proceed to Phase 2"
  )
}
phase2_tbl <- function(w) {
  cm2 <- w * CM2
  tribble(
    ~Step, ~Action,                                                              ~Detail,
    "1",  "Aspirate Cocktail C",                                                 "—",
    "2",  paste0("Accutase: ",fmt(cm2*0.1,0)," µL (100 µL/cm²); 30 min 37°C"), "Full 30 min — do not shorten",
    "3",  paste0("Add NC-C: ",fmt(cm2*0.1,0)," µL; harvest with serological pipette"), "—",
    "4",  "Spin 290×g 1 min RT; resuspend in NC-C",                             "—",
    "5",  paste0("Transfer to ULA plates: 2 mL/well × ",w," wells"),            "Certified ULA only — Corning Costar",
    "6",  "Incubate 2 days",                                                      "—",
    "7",  "Day 14: swirl to center; aspirate from circumference carefully",      "Never aspirate spheroids with media",
    "8",  "Add 2 mL fresh NC-C/well; incubate 24h",                             "Target: 100–300 µm, compact, round"
  )
}
phase3_tbl <- function(w, pp121) {
  cm2 <- w * CM2
  steps <- tribble(
    ~Step, ~Action,                                                                       ~Detail,
    "1",  "Prepare PLO/LM/FN coating on Day 14 (see Surface Coating section)",           "Do not let surface dry",
    "2",  "Swirl ULA plates; remove NC-C from circumference with P1000",                  "—",
    "3",  "Wash with PBS; add Accutase (1 mL/well); 30 min 37°C",                        "—",
    "4",  "⚠ USE 5 mL SEROLOGICAL PIPETTE — 2–3 gentle rounds",                          "Small clusters only; P1000 causes excess cell death",
    "5",  "Spin 290×g 1 min; count viable cells with hemocytometer + trypan blue",       "—",
    "6",  paste0("Plate 100,000 viable cells/cm² — ",fmt(cm2*1e5/1e6,1),"M cells total"),"Move plates N/S/E/W after returning to incubator"
  )
  if (isTRUE(pp121)) {
    steps <- bind_rows(steps, tibble(
      Step   = "7",
      Action = "⚠ Add PP121 (1–2 µM) — NOS1+ enrichment",
      Detail = "Days 15–20 ONLY. Remove from EN-C Day 20. Depletes CHAT+ cholinergic neurons."
    ))
  }
  steps
}
feeding_tbl <- function(w) {
  cm2 <- w * CM2
  tribble(
    ~`Day range`, ~Frequency,        ~`µL/cm²`, ~`Total/feed (µL)`, ~Notes,
    "15–40",      "Every other day", "200",     fmt(cm2*0.2,0),     "Tilt; add from corner — never onto neurites",
    "40–75",      "1–2×/week",       "400",     fmt(cm2*0.4,0),     "Wait for yellow before feeding",
    ">75",        "When yellow",     "400",     fmt(cm2*0.4,0),     "Supplement with FN+LM if detaching"
  )
}
stbl <- function(df, ...) {
  kbl(df, ...) |>
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE, font_size = 13)
}
```

```{r server}
#| context: server
output$summary <- renderUI({
  w   <- input$n_wells
  cm2 <- w * CM2
  tagList(
    tags$p(tags$strong(paste0(w," well(s)  ·  ",cm2," cm² total  ·  ",input$hpsc_line))),
    tags$p("All master mix volumes include +10% excess.")
  )
})
output$ca       <- renderTable({ cocktail_a(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$cb       <- renderTable({ cocktail_b(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$cc       <- renderTable({ cocktail_c(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$ncc      <- renderTable({ ncc_medium(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$enc      <- renderTable({ enc_medium(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$coating  <- renderTable({ coating_tbl(input$n_wells) }, striped=TRUE, bordered=TRUE, spacing="xs")
output$schedule <- renderTable({ day_schedule(input$n_wells) },striped=TRUE, bordered=TRUE, spacing="xs")
output$phase2   <- renderTable({ phase2_tbl(input$n_wells) },  striped=TRUE, bordered=TRUE, spacing="xs")
output$phase3   <- renderTable({ phase3_tbl(input$n_wells, input$pp121) }, striped=TRUE, bordered=TRUE, spacing="xs")
output$feeding  <- renderTable({ feeding_tbl(input$n_wells) }, striped=TRUE, bordered=TRUE, spacing="xs")
output$pp121_callout <- renderUI({
  if (isTRUE(input$pp121)) {
    tags$div(
      class = "callout callout-style-default callout-warning",
      tags$div(class = "callout-header",
        tags$div(class = "callout-icon-container",
          tags$i(class = "callout-icon")),
        tags$div(class = "callout-title-container flex-fill", "PP121 active")),
      tags$div(class = "callout-body-container callout-body",
        tags$p("Add PP121 at 1–2 µM from Day 15 to Day 20 only.",
               " Enriches NOS1+ nitrergic neurons at the expense of CHAT+ cholinergic neurons.",
               " Remove completely from EN-C on Day 20 — extended treatment reduces viability."))
    )
  }
})
```

---

# Configure experiment {#configure-experiment .unnumbered}

```{r inputs}
#| panel: sidebar
numericInput("n_wells",   "6-well plate wells",          value = 6, min = 1, max = 96)
selectInput( "hpsc_line", "hPSC line",
             choices = c("H9 (WA09)","UCSF4","WTC11","Custom"))
checkboxInput("pp121",    "PP121 — NOS1+ enrichment (Days 15–20)", value = FALSE)
hr()
tags$small(
  tags$em("When ready:"), tags$br(),
  tags$strong("File → Print → Save as PDF"), tags$br(),
  tags$em("Sidebar hides automatically.")
)
```

```{r summary-out}
#| panel: fill
uiOutput("summary")
uiOutput("pp121_callout")
```

---

::: {.callout-note}
## Overview
hPSC → vagal ENCC (Days 0–12) → crestospheres (Days 12–15) → 2D ENS culture (Days 15–40+).
Always use **Option B** (E6, chemically defined, xeno-free) — required for downstream chromatin and genomics work.
Full neurochemical maturity by Day 55–75. Cultures lasting >100 days achievable with careful feeding.
:::

---

# Media compositions

::: {.callout-note collapse="true"}
## Math audit — how every volume is calculated

All volumes computed by R from your sidebar inputs. Expand if a number looks wrong.

**Surface area** — the foundation of all ENS volume calculations:

$$A_{total} = N_{wells} \times 9.5 \; cm^2/well$$

9.5 cm² is the growth area of one well in a standard 6-well plate (Corning Costar).
Every volume below derives from this number.

**Cocktail volumes** — seeding density drives the media volume:

$$V_{cocktail} = A_{total} \times V_{per\;cm^2}$$

| Phase | Days | µL/cm² |
|-------|------|--------|
| Cocktails A & B | 0–4 | 200 |
| Cocktail C early | 6–8 | 400 |
| Cocktail C late | 10 | 600 |
| EN-C feeds | 15–40 | 200 |
| EN-C late feeds | 40+ | 400 |

**Factor dilutions** — each growth factor is added as a small volume to E6 base.
For Cocktail A, BMP4 at 1:10,000 dilution means:

$$V_{BMP4} = V_{total} \times 10^{-4}$$

The E6 base volume is then: $V_{E6} = V_{total} - \sum V_{factors}$

**NC-C medium** — scales by number of ULA wells (2 mL per well):

$$V_{NC\text{-}C\;total} = N_{wells} \times 2 \; mL$$

**Coating volumes** (PLO, LM/FN) — 100 µL/cm²:

$$V_{coating} = A_{total} \times 0.1 \; mL/cm^2$$

**Plating density** (Phase 3):

$$N_{cells\;total} = A_{total} \times 100{,}000 \; cells/cm^2$$

If your cell count after dissociation is lower than this, you will get sparse cultures.
If higher, you risk overgrowth and detachment. Adjust your plating area, not your
seeding density.

**Where N\_wells =** your current sidebar input for number of 6-well plate wells.
:::

## Cocktail A — Days 0–2 (200 µL/cm²)

::: {.callout-note collapse="true"}
## Why these signals?
BMP4 (1 ng/mL) activates the NC gene regulatory network in the absence of serum — normally provided by undefined KSR components in Option A. SB431542 blocks TGFβ to prevent mesendodermal fate. CHIR at 600 nM activates WNT to initiate NC specification. Lower CHIR here than in later cocktails — high WNT too early overspecifies posterior identity before NC identity is established.
:::

```{r}
tableOutput("ca")
```

## Cocktail B — Days 2–4 (200 µL/cm²)

::: {.callout-note collapse="true"}
## Why the CHIR increase?
CHIR rises from 600 nM to 1.5 µM at Day 2. Once NC identity is initiated, sustained higher WNT signaling drives proliferation and maintains SOX10+ progenitor state. BMP4 is dropped — it is only needed in the short induction window at Day 0.
:::

```{r}
tableOutput("cb")
```

## Cocktail C — Days 6–12 (vagal patterning + retinoic acid)

::: {.callout-warning}
## Volume increases — do not miss this
- Days 6–8: **400 µL/cm²** (doubles from Cocktail B)
- Day 10: **600 µL/cm²** (increases again)

This compensates for rising confluency. Insufficient volume at this stage causes monolayer detachment and lost cells.
:::

::: {.callout-warning}
## Retinoic acid handling
RA is light-sensitive and degrades rapidly at 37°C. Always make a fresh working stock from your –80°C aliquot on the day of use. Use amber tubes. Do not re-use opened aliquots.
:::

::: {.callout-note collapse="true"}
## Why RA starts at Day 6, not Day 0
Premature RA exposure (Days 0–4) actually inhibits NC induction by over-posteriorizing cells before NC identity is established. RA is only added once CHIR-driven NC specification is underway, then it caudalizes NC cells toward vagal/HOXB3+/HOXB5+ identity required for ENS competence.
:::

```{r}
tableOutput("cc")
```

## NC-C — Crestosphere medium (Days 12–15, 2 mL/ULA well)

::: {.callout-tip}
## FGF2 stability
FGF2 is unstable at 37°C. Add it from a fresh –80°C aliquot to NC-C on the day of use. Carry protein in 0.1% BSA to prevent adsorption to tube walls.
:::

```{r}
tableOutput("ncc")
```

## EN-C — Enteric neuron maturation medium (Days 15+)

::: {.callout-tip}
## After Day 40
Reduce feeding to 1–2×/week but increase volume to 400 µL/cm². Wait until medium turns yellow before feeding — over-feeding mature cultures stresses neurons. If cultures begin detaching, supplement EN-C with FN (2 µg/mL) + LM (2 µg/mL).
:::

```{r}
tableOutput("enc")
```

---

# Surface coating — PLO / Laminin / Fibronectin

::: {.callout-warning}
## Prepare on Day 14 — not Day 15
The PLO overnight step means coating must start the day before plating. If you prepare it Day 15 morning you will lose hours of LM/FN incubation time and get poor cell attachment.
:::

::: {.callout-warning}
## Never let the coated surface dry
Aspirate the LM/FN solution and immediately add cells — no gap. A dried laminin surface is denatured and cells will not adhere.
:::

```{r}
tableOutput("coating")
```

---

# Phase 1 — ENCC induction (Days –2 to 12)

## Day –2: Replate hPSCs

::: {.callout-important}
## Day 0 confluency is the #1 determinant of induction efficiency
Target: **~70% confluent monolayer** on Day 0. Too sparse (<60%) or too dense (>85%) both significantly reduce CD49D+ yield at Day 12. If your cell line tends to grow slowly, plate at 5:4 instead of 5:6.
:::

| Step | Action | Detail |
|------|--------|--------|
| 1 | Aspirate E8-C; EDTA (500 µM in PBS) **3–5 min** | Longer than standard passaging — increases cell separation |
| 2 | Passage at **5:6 ratio** (5 wells → 6 new wells) | — |
| 3 | Plate on Matrigel or Vitronectin XF in E8-C | Vitronectin preferred for fully defined system |
| 4 | Target: ~70% confluent monolayer on Day 0 | Feed every other day with E8-C |

## Days 0–12: Medium changes

::: {.callout-tip}
## Adding medium without detaching cells
As the monolayer gets dense (especially Days 8–12), always tilt the plate and pipette slowly from the lower corner or side. Never pipette directly downward onto confluent cells — shear force detaches them.
:::

```{r}
tableOutput("schedule")
```

### Day 11–12 QC — do not proceed without passing {.unnumbered}

::: {.callout-important}
## QC gate — required before Phase 2
- [ ] FACS: CD49D+ ≥ 50% (target 60–80%) — observed: _______ %
- [ ] qPCR: HOXB3 ≥ 10-fold over Day 0 hPSC — observed: _______ fold
- [ ] qPCR: HOXB5 ≥ 10-fold over Day 0 hPSC — observed: _______ fold
- [ ] Phase contrast: loose, elongated cells; disrupted colony borders

If CD49D+ < 40%, check RA freshness, CHIR lot, and Day 0 confluency before troubleshooting further.
:::

::: {.callout-tip}
## Optional: FACS purify at Day 12 for early progenitor experiments
Sort for CD49D+ using Pe/Cy7 anti-human CD49D (Biolegend 304314). Recommended if you need pure ENC progenitors for Day 11 assays. For most applications, the sphere phase (Days 12–15) provides sufficient enrichment without sorting.
:::

---

# Phase 2 — Crestosphere formation (Days 12–15)

::: {.callout-warning}
## Ultra-low attachment plates — must be certified ULA
Standard tissue culture plates will cause cells to adhere instead of forming spheres. Use Corning Costar certified ultra-low attachment plates only. This is the most common cause of Phase 2 failure.
:::

```{r}
tableOutput("phase2")
```

::: {.callout-tip}
## Sphere size troubleshooting
Spheres should be 100–300 µm at Day 14, compact and round with smooth borders. If spheres have fused into large irregular bodies, gently pipette 2–3× with a P1000 to separate. Necrotic cores develop in large bodies and reduce Day 15 viability.
:::

::: {.callout-important}
## Day 15 QC
- [ ] FACS: CD49D+ ≥ 80–90% — observed: _______ %

If CD49D+ is still <70% after the sphere phase, extend suspension culture to Day 16 before plating.
:::

---

# Phase 3 — 2D ENS differentiation (Days 15–40+)

::: {.callout-warning}
## Serological pipette — not P1000 — for Day 15 dissociation
The narrower P1000 tip adds shear stress that causes excessive cell death. Use a 5 mL serological pipette for gentle dissociation into small clusters (5–20 cells). Single cell suspensions give very poor neuronal yields.
:::

```{r}
tableOutput("phase3")
```

::: {.callout-note collapse="true"}
## Why move plates N/S/E/W after plating?
Cells in suspension in EN-C distribute unevenly as soon as the plate is set flat. A gentle 4-direction rocking motion redistributes them before they settle, giving even coverage across the well. Do this within 2 minutes of placing in the incubator.
:::

## Feeding schedule

```{r}
tableOutput("feeding")
```

::: {.callout-tip}
## Long-term culture (Day 75+)
For cultures older than 50 days, wait until medium turns yellow before feeding — do not feed on a fixed schedule. Use FN/LM-supplemented EN-C at every feed. Cultures lasting >100 days are achievable. The main enemy of long-term cultures is aggressive medium changes that mechanically detach neurites.
:::

---

# QC checkpoints summary

```{r qc-tbl}
qc <- tribble(
  ~Day,    ~Checkpoint,                    ~Expected,                    ~Observed, ~Pass,
  "0",     "hPSC confluency",              "~70% monolayer",              "",        "",
  "6",     "SOX10 clustering",             "GFP+ clusters visible",       "",        "",
  "12",    "ENCC induction (FACS)",        "CD49D+ ≥50%",                "",        "",
  "12",    "Vagal identity (qPCR)",        "HOXB3/5 ≥10-fold",          "",        "",
  "15",    "Sphere enrichment (FACS)",     "CD49D+ ≥80–90%",             "",        "",
  "15",    "Sphere morphology",            "100–300 µm, compact, round",  "",        "",
  "30",    "Early neurons (IF)",           "TUBB3+/TrkC+ networks",       "",        "",
  "40",    "Neuronal purity (FACS)",       ">80% TUBB3+",                "",        "",
  "40",    "Neurochemical (IF)",           "NOS1, CHAT, 5HT, GABA",      "",        "",
  "75",    "Full FACS",                    "CHAT~40%, NOS1~9%, 5HT~6%",  "",        ""
)
stbl(qc)
```

::: {.callout-note collapse="true"}
## Expected neurochemical composition at Day 75
Per Barber et al. 2019 Fig. 4g and Fig. 5b: ~40% CHAT+ (cholinergic), ~9% GABA+, ~9% NOS1+ (nitrergic), ~6% 5HT+ (serotonergic), and ~25% SOX10+/GFAP+ (enteric glia). With PP121 treatment Days 15–20, NOS1+ increases 3–5× at the expense of CHAT+ (Majd et al. 2025).
:::

**Notes / deviations:**

---

# Troubleshooting

::: {.callout-tip collapse="true"}
## CD49D+ <40% at Day 12
Most likely causes: wrong Day 0 confluency; degraded RA; CHIR lot variability. Check in this order: (1) confirm Day 0 was 70%; (2) make fresh RA from a new –80°C aliquot; (3) titrate your CHIR lot (1–5 µM) in a test well. CHIR activity varies significantly between lots.
:::

::: {.callout-tip collapse="true"}
## Spheroids not forming (Phase 2)
Almost always a plate issue — verify Corning ULA certification. If using correct plates, check FGF2 activity (add BSA carrier) and reduce seeding density.
:::

::: {.callout-tip collapse="true"}
## Low TUBB3+ at Day 40 (<60%)
Three most common causes: (1) single-cell dissociation at Day 15 — use serological pipette; (2) PLO/LM/FN surface dried before plating; (3) inactive GDNF lot. Test each in order.
:::

::: {.callout-tip collapse="true"}
## SMA+ myofibroblast contamination
NC-derived SMA+ flat cells are the primary contaminant. They eventually cause neuronal detachment. Minimize by FACS-sorting CD49D+ at Day 12 if contamination is severe, or by reducing passaging ratio at Day –2.
:::

---

# References

Barber K, Studer L, Fattahi F. *Nat Protoc.* 2019;14(4):1261–1279.

Majd H, Samuel RM, et al. *Nature.* 2025;645:158–167.

Fattahi F, et al. *Nature.* 2016;531:105–109.
