---
title: "CUT&RUN — [antibody] — [cell type]"
date: today
author: "JP Flores"
categories: [CUT&RUN, chromatin]
description: ""  # fill in one-line result summary when done
status: "in-progress"  # change to "complete" when committed
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)

stbl <- function(df, ...) {
  kbl(df, ...) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 13)
}
```

## Overview

```{r exp-meta}
#| echo: false
meta <- tribble(
  ~Field,               ~Value,
  "Date",               as.character(Sys.Date()),
  "Cell line / type",   "",
  "Treatment",          "",
  "N samples",          "",
  "Antibodies",         "",
  "Cells per sample",   "",
  "PCR cycles",         "",
  "Aim this supports",  "",
  "Protocol version",   "[protocols/CnR_protocol.qmd](../protocols/_shiny/CnR_protocol.qmd)",
  "Raw data path",      "",
  "Sequencing run ID",  ""
)
stbl(meta, col.names = c("", ""))
```

## Hypothesis

<!-- What do you expect to see and why? -->

## Antibody panel

```{r antibody-table}
abs <- tribble(
  ~Sample, ~Antibody, ~Lot, ~Dilution, ~`Expected result`,
  "1",     "H3K4me3 (positive ctrl)", "", "1:100", "Clear peaks at active promoters",
  "2",     "IgG (negative ctrl)",     "", "1:100", "Flat / background only",
  "3",     "",                         "", "",       "",
  "4",     "",                         "", "",       ""
)
stbl(abs, caption = "Antibody panel — fill in before Day 1")
```

---

## Day 1 — Cell binding to ConA beads {.day-header}

**Date:** ___ **Started:** ___ **Finished:** ___

::: {.callout-important appearance="minimal"}
Always include H3K4me3 positive control and IgG negative control in every run.
:::

```{r day1-qc}
qc1 <- tribble(
  ~Checkpoint,                    ~Expected,                ~Observed, ~Pass,
  "Cell count (hemocytometer)",   "~500k/sample",           "",        "",
  "Viability (trypan blue)",      ">85%",                   "",        "",
  "Bead binding appearance",      "Even suspension, no clumps","",     "",
  "Antibody added + recorded",    "Yes",                    "",        "",
  "In thermomixer at 4°C O/N",   "Yes — 1000 rpm",         "",        ""
)
stbl(qc1, caption = "Day 1 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 2 — pAG-MNase & chromatin digestion {.day-header}

**Date:** ___ **CaCl₂ added at:** ___ **Digest completed at:** ___

::: {.callout-warning appearance="minimal"}
CaCl₂ triggers digestion immediately — work consistently across all samples.
Digest time: exactly 30 min at 4°C. Supernatant = your fragments.
:::

```{r day2-qc}
qc2 <- tribble(
  ~Checkpoint,                    ~Expected,              ~Observed, ~Pass,
  "Digest duration",              "30 min 4°C",           "",        "",
  "37°C stop incubation",         "20 min done",          "",        "",
  "Supernatant appearance",       "Clear (fragments released)","",   "",
  "Post-Zymo Qubit HS (ng/µL)",  ">1",                   "",        ""
)
stbl(qc2, caption = "Day 2 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 3 — Library preparation {.day-header}

**Date:** ___ **PCR cycles:** ___

::: {.callout-warning appearance="minimal"}
Record adapter assignments before opening any tubes — index swaps are invisible until sequencing.
:::

```{r day3-adapters}
adapters <- tribble(
  ~Sample, ~Antibody, ~Index, ~`Post-ligation Qubit (ng/µL)`, ~`Final Qubit (ng/µL)`, ~`Tapestation peak (bp)`,
  "1",     "H3K4me3", "",    "",                               "",                      "",
  "2",     "IgG",     "",    "",                               "",                      "",
  "3",     "",        "",    "",                               "",                      "",
  "4",     "",        "",    "",                               "",                      ""
)
stbl(adapters, caption = "Adapter assignments and final QC")
```

```{r day3-qc}
qc3 <- tribble(
  ~Checkpoint,                  ~Expected,      ~Observed, ~Pass,
  "Post-ligation Qubit (ng/µL)","≥5",           "",        "",
  "Final library Qubit (ng/µL)","≥2",           "",        "",
  "Tapestation peak (bp)",      "150–350",      "",        "",
  "Submitted for sequencing",   "Yes",          "",        ""
)
stbl(qc3, caption = "Day 3 QC")
```

**Sequencing run ID / submission date:**

**Deviations from protocol:**

**Notes:**

---

## Summary QC table

```{r summary-qc}
qc_summary <- tribble(
  ~Checkpoint,                    ~Expected,     ~Observed, ~Pass,
  "Cell count per sample",        "~500k",       "",        "",
  "Viability",                    ">85%",        "",        "",
  "Post-Zymo Qubit (ng/µL)",     ">1",          "",        "",
  "Post-ligation Qubit (ng/µL)", ">5",          "",        "",
  "Final library Qubit (ng/µL)", ">2",          "",        "",
  "Tapestation peak (bp)",        "150–350",     "",        ""
)
stbl(qc_summary, caption = "Full experiment QC summary")
```

---

## Interpretation

<!-- What did you see? Positive and negative controls behaving? Peaks where expected? -->

## Deviations summary

<!-- Consolidate all deviations here -->

## Next steps

## References

- Skene PJ, Henikoff JG, Henikoff S. *eLife.* 2018;7:e36906.
- Protocol: [protocols/CnR_protocol.qmd](../protocols/_shiny/CnR_protocol.qmd)
