---
title: "Hi-C — [target] — [cell type]"
date: today
author: "JP Flores"
categories: [Hi-C, chromatin]
description: ""  # fill in one-line result summary when done — shows in experiments/index
status: "in-progress"  # change to "complete" when committed
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)

stbl <- function(df, ...) {
  kbl(df, ...) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 13)
}
```

## Overview

```{r exp-meta}
#| echo: false
meta <- tribble(
  ~Field,              ~Value,
  "Date started",      as.character(Sys.Date()),
  "Date completed",    "",
  "Cell line / type",  "",
  "Treatment",         "",
  "N samples",         "",
  "Aim this supports", "",
  "Protocol version",  "[protocols/HiC_protocol.qmd](../protocols/_shiny/HiC_protocol.qmd)",
  "Raw data path",     "",
  "Sequencing run ID", ""
)
stbl(meta, col.names = c("", ""))
```

## Hypothesis

<!-- What do you expect to see and why? One paragraph. -->

## Experimental design

<!-- Which samples? What's the comparison? Any controls? -->

---

## Day 0 — Crosslinking {.day-header}

**Date:** ___ **Started:** ___ **Finished:** ___

```{r day0-qc}
qc0 <- tribble(
  ~Checkpoint,                      ~Expected,                ~Observed, ~Pass,
  "Cell count (hemocytometer)",     "~5M/sample",              "",        "",
  "Viability (trypan blue)",        ">85%",                    "",        "",
  "Pellet appearance post-fixation","Firm, off-white",         "",        "",
  "Formaldehyde lot #",             "",                        "",        ""
)
stbl(qc0, caption = "Day 0 QC")
```

::: {.callout-note appearance="minimal" collapse="true"}
## Protocol reminder — Day 0
Formaldehyde 1% final (16% stock), 10 min RT. Quench with 2M glycine to 0.2M. Wash 2× cold PBS. Flash freeze in LN₂.
:::

**Deviations from protocol:**

**Notes:**

---

## Day 1 — Lysis & MboI restriction digest {.day-header}

**Date:** ___ **Started:** ___ **MboI lot #:** ___

```{r day1-qc}
qc1 <- tribble(
  ~Checkpoint,              ~Expected,           ~Observed, ~Pass,
  "SDS permeabilization",   "62°C 7 min done",   "",        "",
  "Triton quench",          "37°C 15 min done",  "",        "",
  "MboI digest duration",   "O/N or ≥2 hr",      "",        "",
  "MboI returned to –20°C", "Yes",               "",        ""
)
stbl(qc1, caption = "Day 1 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 2 — Biotin fill-in, ligation & crosslink reversal {.day-header}

**Date:** ___ **Biotin fill-in started:** ___ **Ligation started:** ___

::: {.callout-warning appearance="minimal"}
Biotin fill-in: 5% excess only. Ligation: RT slow rotation, door open.
:::

```{r day2-qc}
qc2 <- tribble(
  ~Checkpoint,                ~Expected,            ~Observed, ~Pass,
  "MboI inactivation",        "62°C 20 min done",   "",        "",
  "Biotin fill-in",           "37°C 1.5h done",     "",        "",
  "Ligation duration",        "4h RT",               "",        "",
  "Crosslink reversal started","55°C 30 min → 68°C O/N","",   ""
)
stbl(qc2, caption = "Day 2 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 3 — DNA precipitation {.day-header}

**Date:** ___ **–80°C precipitation started:** ___

```{r day3-qc}
qc3 <- tribble(
  ~Checkpoint,              ~Expected,          ~Observed, ~Pass,
  "Pellet visible post-spin","Yes",              "",        "",
  "Qubit BR dsDNA",         "measurable",        "",        "",
  "Pellet dissolved in Tris","37°C 1h done",     "",        ""
)
stbl(qc3, caption = "Day 3 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 4 — Shearing & size selection {.day-header}

**Date:** ___ **Covaris settings:** Duty 25 / PIP 500 / 200 cyc/burst / 90 sec

```{r day4-qc}
qc4 <- tribble(
  ~Checkpoint,                       ~Expected,          ~Observed, ~Pass,
  "Pre-shear Qubit BR (ng/µL)",      ">10",              "",        "",
  "Post-shear gel",                  "smear 200–2000 bp", "",       "",
  "Post-size-selection Qubit HS",    ">5 ng/µL",         "",        "",
  "Post-size-selection gel",         "300–500 bp",       "",        ""
)
stbl(qc4, caption = "Day 4 QC")
```

**Deviations from protocol:**

**Notes:**

---

## Day 5 — Biotin pull-down & library prep {.day-header}

**Date:** ___

```{r day5-qc}
qc5 <- tribble(
  ~Sample, ~Index, ~Notes,
  "",      "",     "",
  "",      "",     "",
  "",      "",     ""
)
stbl(qc5, caption = "Adapter index assignments — fill in before opening tubes")
```

**Deviations from protocol:**

**Notes:**

---

## Day 6 — PCR amplification & final cleanup {.day-header}

**Date:** ___ **PCR cycles used:** ___

```{r day6-qc}
qc6 <- tribble(
  ~Sample, ~`Qubit HS (ng/µL)`, ~`Tapestation peak (bp)`, ~`Submitted?`,
  "",      "",                   "",                        "",
  "",      "",                   "",                        "",
  "",      "",                   "",                        ""
)
stbl(qc6, caption = "Final library QC")
```

**Sequencing run ID / submission date:**

**Deviations from protocol:**

**Notes:**

---

## Summary QC table

```{r summary-qc}
qc_summary <- tribble(
  ~Checkpoint,                        ~Expected,           ~Observed, ~Pass,
  "Pre-shear Qubit BR (ng/µL)",       ">10",               "",        "",
  "Post-shear gel",                   "smear 200–2000 bp", "",        "",
  "Post-size-selection Qubit (ng/µL)","≥5",                "",        "",
  "Post-size-selection gel",          "300–500 bp",        "",        "",
  "Final library Qubit (ng/µL)",      ">2",                "",        "",
  "Tapestation peak (bp)",            "300–600",           "",        ""
)
stbl(qc_summary, caption = "Full experiment QC summary")
```

---

## Interpretation

<!-- What did you see? Does it match your hypothesis? -->

## Deviations summary

<!-- Consolidate all deviations from the 6 days here for easy reference -->

## Next steps

<!-- What does this experiment tell you to do next? -->

## References

- Rao SSP et al. *Cell.* 2014;159(7):1665–1680.
- Protocol: [protocols/HiC_protocol.qmd](../protocols/_shiny/HiC_protocol.qmd)
